Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-02-16 numerics_paper/version2/numerical_chj.tex
more changes

\subsubsection{Regime I}
In this case, the boundary conditions are given by \eqref{bcs_chj} with
\begin{align}\label{bcs_chj_r1}
  |\bfu_I|=0.75, \qquad h_o=0.37387387318873766. 
\end{align}
The initial condition is a symmetric CHJ located at $\sqrt{x^2+y^2}=0.3$, generated as described in \S \ref{sec:steady_chj}.
In Figure \ref{fig:chj_r1_roe}, we show the solution at different times using
method \eqref{second-order_via_fluct} with Roe's Riemann solver.
The solution clearly develops carbuncle instabilities.
In Figure \ref{fig:chj_r1_diff}, we show solutions at $t=3$ for
Rusanov's, HLLEMCC, and the blended Riemann solvers.
For this test case, these three solvers give qualitatively similar solutions,
all of which are free from carbuncles.
%however, we expect Rusanov's solver to be less accurate.

\begin{figure}[!h]
  \centering
  \subfloat[Solution at (left) $t=1$, (middle) $t=1.5$ and (right) $t=3$ using Roe's Riemann solver.\label{fig:chj_r1_roe}]{
    \begin{tabular}{ccc} 
      \includegraphics[scale=0.38]{figures/chj_r1_roe_p1.png} &
      \includegraphics[scale=0.38]{figures/chj_r1_roe_p2.png} &
      \includegraphics[scale=0.38]{figures/chj_r1_roe_p3.png}
    \end{tabular}
  }

  \subfloat[Solution at $t=3$ using (left) Rusanov's, (middle) HLLEMCC and (right) the blended solvers.\label{fig:chj_r1_diff}]{
    \begin{tabular}{ccc}
      \includegraphics[scale=0.38]{figures/chj_r1_llf_p3.png} &
      \includegraphics[scale=0.38]{figures/chj_r1_hllemcc_p3.png} &
      \includegraphics[scale=0.38]{figures/chj_r1_blended_p3.png}
    \end{tabular}
    }
    \caption{CHJ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r1}.    
      We consider method \eqref{second-order_via_fluct} with different Riemann solvers.
      In all cases we show the Schlieren plot for the water height $h$.
      In (a), for the plot that corresponds to $t=3$, we zoom close to the jump
      and plot the y-momentum. 
    \label{fig:chj_r1}}
\end{figure}

\subsubsection{Regime II}
We now consider a more aggresive regime. The boundary conditions are given by \eqref{bcs_chj} with 
\begin{align}\label{bcs_chj_r2}
  |\bfu_I|=15, \qquad h_o=6.6845019298155357.
\end{align}
Again we follow \S \ref{sec:steady_chj} to generate an initial condition with a symmetric CHJ located at
$\sqrt{x^2+y^2}=0.3$. In Figure \ref{fig:chj_r2_roe}, we show the solution at different times using method
\eqref{second-order_via_fluct} with Roe's Riemann solver. Again the solution develops
carbuncle instabilities. For this regime, we obtained negative values for the water height with
the HLLEMCC solver, which lead to failure of the solver. Therefore, we only show results with
Rusanov's and the blended solvers, see Figure \ref{fig:chj_r2_diff}. With both solvers, the carbuncles are avoided.
%
In contrast to what we observed in the first test case, in this situation the Froude number is
high enough that numerical perturbations start to trigger a response around the CHJ
(see the middle and right panels of Figure \ref{fig:chj_r2_diff}).
For the resolution that we consider, this behavior is dissipated by the extra numerical viscosity introduced by
Rusanov's solver (see the left panel of Figure \ref{fig:chj_r2_diff}).
In Figure \ref{fig:chj_r2_later}, we show the solution at a later time considering the same solvers as before.
This time, we plot not only the Schlieren plot for the water height, but also the magnitude of the momentum. 
Clearly, using Rusanov's solver dissipates the physical response to these numerical perturbations.

@@

  \vspace{10pt}
  \subfloat[Solution at $t=0.1$ using (left) Rusanov's and (right) the blended solvers.\label{fig:chj_r2_diff}]{
    \begin{tabular}{ccc}
      \includegraphics[scale=0.3]{figures/chj_r2_llf_t0p1.png} \qquad &
      %\quad \includegraphics[scale=0.5]{figures/cross.png} \quad &
      \qquad \includegraphics[scale=0.3]{figures/chj_r2_blended_t0p1.png}
    \end{tabular}
    }
  \caption{CHJ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r2}. 
    We consider method \eqref{second-order_via_fluct} with different Riemann solvers.
    In all cases we show the Schlieren plot for the water height $h$.
    %The color scale in the plots go from (blue) $0.45$ to (red) $4.5$,
    %which corresponds to the minimum and maximum values of the momentum at $t=0$.
  }
\end{figure}

\begin{figure}[!h]
  \centering
  \subfloat[$t=5$]{
    \includegraphics[scale=0.3]{figures/chj_r2_rusanov_t5p0.png} \quad
    \includegraphics[scale=0.5]{figures/cross.png} \quad
    \includegraphics[scale=0.3]{figures/chj_r2_blended_t5p0.png}
  }
  \caption{CHJ at $t=5$ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r2}. 
    We consider method \eqref{second-order_via_fluct} with (left) Rusanov's,
    (middle) HLLEMCC and (right) the blended solvers. 
    In all cases we superimpose the Schlieren plot for the water height $h$
    an the magnitude of the momentum with a uniform (white-to-blue) colormap. 
    The scale for colormap is (white) $0.45$ to (blue) $4.5$, 
    which corresponds to the minimum and maximum values at $t=0$.
    \label{fig:chj_r2_later}}
\end{figure}

\subsubsection{Regime II with random perturbations}
Finally, we introduce a physical random perturbation to the inflow boundary condition.
We use the same mean values \eqref{bcs_chj_r2}, but now we set
\begin{subequations}
  \begin{align}
    h(x,y,t)=\tilde h_I, \quad
    u(x,y,t)=|\tilde \bfu_I| \left(\frac{x}{r_I}\right), \quad
    v(x,y,t)=|\tilde \bfu_I| \left(\frac{y}{r_I}\right), \qquad \forall (x,y)\in\partial\Omega_I,
  \end{align}
\end{subequations}
with $\tilde h_I = \frac{h_I}{1+\epsilon(x,y,t)}$ and $|\tilde\bfu_I|=|\bfu_I|(1+\epsilon(x,y,t))$.
Here $\epsilon(x,y,t)$ is chosen at each inflow boundary ghost cell and at each
time step as an i.i.d. random variable from the uniform distribution $[-0.01,0.01]$.
In this case, due to the large Froude number, the random perturbations in the inflow
may trigger physical instabilities, which we want to preserve. At the same time, we
do not want the solution to develop carbuncle instabilities.
In figures \ref{fig:chj_r2wp_rusanov} and \ref{fig:chj_r2wp_blended}, we show the solution
at different times considering method \eqref{second-order_via_fluct} with Rusanov's and
the blended Riemann solvers, respectively.

@@

  \subfloat[Blended Riemann solver\label{fig:chj_r2wp_blended}]{
    \begin{tabular}{cccc}
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t2p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t3p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t4p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t5p0.png}
    \end{tabular}
  }
  \caption{CHJ at different times with a random perturbation at the inflow boundary.
    We consider method \eqref{second-order_via_fluct} with (a) Roe's and (b) the blended
    Riemann solvers. In all cases we show the Schlieren plot for the water height $h$.}
\end{figure}

>>>>

\subsubsection{Regime I}
In this case, the boundary conditions are given by \eqref{bcs_chj} with
\begin{align}\label{bcs_chj_r1}
  |\bfu_I|=0.75, \qquad h_o=0.37387387318873766. 
\end{align}
The initial condition is a symmetric CHJ located at $\sqrt{x^2+y^2}=0.3$, generated as described in \S \ref{sec:steady_chj}.
In Figure \ref{fig:chj_r1_roe}, we show the solution at different times using
method \eqref{second-order_via_fluct} with Roe's Riemann solver.
The solution clearly develops carbuncle instabilities, which are evident in the inset figure at $t=3$.
In Figure \ref{fig:chj_r1_diff}, we show solutions at $t=3$ for
Rusanov's, HLLEMCC, and the blended Riemann solvers.
For this test case, these three solvers give qualitatively similar solutions,
all of which are free from carbuncles.
%however, we expect Rusanov's solver to be less accurate.

\begin{figure}[!h]
  \centering
  \subfloat[Solution at (left) $t=1$, (middle) $t=1.5$ and (right) $t=3$ using Roe's Riemann solver.\label{fig:chj_r1_roe}]{
    \begin{tabular}{ccc} 
      \includegraphics[scale=0.3]{figures/chj_r1_roe_p1.png} &
      \includegraphics[scale=0.3]{figures/chj_r1_roe_p2.png} &
      \includegraphics[scale=0.3]{figures/chj_r1_roe_p3.png}
    \end{tabular}
  }

  \subfloat[Solution at $t=3$ using (left) Rusanov's, (middle) HLLEMCC and (right) the blended solvers.\label{fig:chj_r1_diff}]{
    \begin{tabular}{ccc}
      \includegraphics[scale=0.3]{figures/chj_r1_llf_p3.png} &
      \includegraphics[scale=0.3]{figures/chj_r1_hllemcc_p3.png} &
      \includegraphics[scale=0.3]{figures/chj_r1_blended_p3.png}
    \end{tabular}
    }
    \caption{CHJ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r1}.    
      We consider method \eqref{second-order_via_fluct} with different Riemann solvers.
      In all cases we show the Schlieren plot for the water height $h$.
      In (a), for the plot that corresponds to $t=3$, we zoom close to the jump
      and plot the y-momentum. 
    \label{fig:chj_r1}}
\end{figure}

\subsubsection{Regime II}
We now consider a more aggresive regime. The boundary conditions are given by \eqref{bcs_chj} with 
\begin{align}\label{bcs_chj_r2}
  |\bfu_I|=15, \qquad h_o=6.6845019298155357.
\end{align}
Again we follow \S \ref{sec:steady_chj} to generate an initial condition with a symmetric CHJ located at
$\sqrt{x^2+y^2}=0.3$. In Figure \ref{fig:chj_r2_roe}, we show the solution at different times using method
\eqref{second-order_via_fluct} with Roe's Riemann solver. Again the solution develops
carbuncle instabilities, which are clearly seen in the inset figure at $t=3$.
For this regime, we obtained negative values for the water height with
the HLLEMCC solver, which lead to failure of the solver. Therefore, we only show results with
Rusanov's and the blended solvers, see Figure \ref{fig:chj_r2_diff}. With both solvers, the carbuncles are avoided.
%
In contrast to what we observed in the first test case, in this situation the Froude number is
high enough that numerical perturbations start to trigger a response around the CHJ
(see the right panel of Figure \ref{fig:chj_r2_diff}).
For the resolution that we consider, this behavior is dissipated by the extra numerical viscosity introduced by
Rusanov's solver (see the left panel of Figure \ref{fig:chj_r2_diff}).
In Figure \ref{fig:chj_r2_later}, we show the solution at a later time considering the same solvers as before.
This time, we plot not only the Schlieren plot for the water height, but also the magnitude of the momentum. 
Clearly, using Rusanov's solver dissipates the physical response to these numerical perturbations.

@@

  \vspace{10pt}
  \subfloat[Solution at $t=0.1$ using (left) Rusanov's and (right) the blended solvers.\label{fig:chj_r2_diff}]{
    \begin{tabular}{ccc}
      \includegraphics[scale=0.3]{figures/chj_r2_llf_t0p1.png} \qquad &
      %\quad \includegraphics[scale=0.5]{figures/cross.png} \quad &
      \qquad \includegraphics[scale=0.3]{figures/chj_r2_blended_t0p1.png}
    \end{tabular}
    }
  \caption{CHJ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r2}. 
    We consider method \eqref{second-order_via_fluct} with different Riemann solvers.
    In all cases we show the Schlieren plot for the water height $h$.
    In (a), for the plot that corresponds to $t=0.11$, we zoom close to the jump and plot the y-momentum.}
\end{figure}

\begin{figure}[!h]
  \centering
    \includegraphics[scale=0.3]{figures/chj_r2_rusanov_t5p0.png} \quad
    %\includegraphics[scale=0.5]{figures/cross.png} \quad
    \quad \includegraphics[scale=0.3]{figures/chj_r2_blended_t5p0.png}
  \caption{CHJ at $t=5$ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r2}. 
    We consider method \eqref{second-order_via_fluct} with (left) Rusanov's
    and (right) the blended solvers. 
    In all cases we superimpose the Schlieren plot for the water height $h$
    an the magnitude of the momentum with a uniform (white-to-blue) colormap. 
    The scale for colormap is (white) $0.45$ to (blue) $4.5$, 
    which corresponds to the minimum and maximum values at $t=0$.
    \label{fig:chj_r2_later}}
\end{figure}

\subsubsection{Regime II with random perturbations}
Finally, we introduce a physical random perturbation to the inflow boundary condition.
We use the same mean values \eqref{bcs_chj_r2}, but now we set
\begin{subequations}
  \begin{align}
    h(x,y,t)=\tilde h_I, \quad
    u(x,y,t)=|\tilde \bfu_I| \left(\frac{x}{r_I}\right), \quad
    v(x,y,t)=|\tilde \bfu_I| \left(\frac{y}{r_I}\right), \qquad \forall (x,y)\in\partial\Omega_I,
  \end{align}
\end{subequations}
with $\tilde h_I = \frac{h_I}{1+\epsilon(x,y,t)}$ and $|\tilde\bfu_I|=|\bfu_I|(1+\epsilon(x,y,t))$.
Here $\epsilon(x,y,t)$ is chosen at each inflow boundary ghost cell and at each
time step as an i.i.d. random variable from the uniform distribution $[-0.01,0.01]$.
In this case, due to the large Froude number, the random perturbations in the inflow
may trigger physical instabilities, which we want to preserve. At the same time, we
do not want the solution to develop carbuncle instabilities.
In figures \ref{fig:chj_r2wp_rusanov} and \ref{fig:chj_r2wp_blended}, we show the solution
at different times considering method \eqref{second-order_via_fluct} with Rusanov's and
the blended Riemann solvers, respectively.
In the right panel of Figure \ref{fig:chj_r2wp_blended}, we show an inset figure 
with the y-momentum. Note that even though the solution is highly unstable 
(with the blended Riemann solver), no visible carbuncle instabilities are developed.

@@

  \subfloat[Blended Riemann solver\label{fig:chj_r2wp_blended}]{
    \begin{tabular}{cccc}
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t2p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t3p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t4p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t5p0.png}
    \end{tabular}
  }
  \caption{CHJ at different times with a random perturbation at the inflow boundary.
    We consider method \eqref{second-order_via_fluct} with different Riemann solvers. 
    In all cases we show the Schlieren plot for the water height $h$.
    In (b), for the plot that corresponds to $t=5$, we zoom close to the jump and plot the y-momentum.}
\end{figure}
